Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest

### 2022-07-20 article/disc.Rmd
draft disc

---
title: "Discussion"
output: 
  # html_notebook
  bookdown::html_document2:
    # number_sections: FALSE
bibliography: ../../library.bib
link-citations: TRUE
---

@@

This study found significant edge effects on secondary forest stand wood and stem density, tree diversity and community composition, and some functional traits, yet not on overall aboveground biomass.
<!-- These results were shown in Figs \@ref(fig:biomass), \@ref(fig:taxa), and \@ref(fig:traits)... -->
It was initially hypothesized that community composition would vary with distance to edge, favoring more shade-tolerant taxa, due to light availability and dispersal potential, and this evidence supported edge effects on community composition and biomass of taxa associated with both early- and late-successional stages, but without detectable roles for light availability or general dispersal mode.
Accordingly this suggests that edge effects do significantly change humid tropical forest community composition, but primarily due to factors other than light competition or dispersal limitation, as well as that aboveground biomass and even low taxonomic richness is resilient after a decade of regeneration.

Aboveground biomass storage is a key global ecosystem function and service, but edge effects tend to release about one-third of carbon lost from tropical deforestation *[@brinck17]*, with more expected in the future *[@mitchard18]*.
<!-- responds variably to secondary succession and edge effects ...   -->
Management has been found to explain most of biomass variation among tropical forest fragments, and wood density ~10% biomass variation *[@pyles22]*.
This study adds that management via edge effects can also mediate ~10% of wood density variation without affecting local biomass, although median stem wood density can describe regional spatial biomass patterns *[@baker04]*.
Overall, wood density and richness values may have been pulled down by early-successional clonal trees closest to the primary forest, which tend to have lower wood densities, lowering carbon *[@gonzalezj98; @pyles22]*, and increasing nearby stem density.
Lower wood density can also reflect more dynamic forest stands *[@malhi06]* with higher tree turnover, a consequence of forest fragmentation *[@nascimento04a]* together with other global change drivers *[@laurance14]*, as well as may reflect possible differences in soil fertility *[@malhi04]*, all of which represent possible future research directions, beyond strictly neutral dynamics *[@terborgh95]*.
Furthermore, light availability at the plot level may not affect biomass storage, but instead individual light gaps may be more important for biomass dynamics *[@chazdon84]*.
    <!-- - although canopy light may still affect biomass generally in this case, without edge effects (check direct light-biomass correlation graph? -- no sig correlation) -->

Edge effects have been shown to lower both forest biomass and also diversity, and this study adds that the decline can be rapid and non-linear across edge gradients, along with supporting the likelihood of lasting changes to community composition.
The rapid diversity loss across the edge gradient may be in part due to finer variations in dispersal patterns among animal taxa, rather than overall dispersal mode more broadly.
Similarly, stochastic factors like priority effects after dispersal may also affect offspring establishment and re-growth.
While the biomass of the most abundant taxon *Vochysia* was resilient to edge effects, that of the key taxon for frugivores like birds and bats *Ficus* was significantly lowered across the forest edge, pointing to potentially different responses of biomass and food web associated ecosystem services.

>>>>

---
title: "Discussion"
output: 
  html_notebook
  # bookdown::html_document2:
    # number_sections: FALSE
bibliography: ../../library.bib
link-citations: TRUE
---

@@

This study found significant edge effects on secondary forest stand wood and stem density, tree diversity and community composition, and some functional traits, yet not on overall aboveground biomass.
<!-- These results were shown in Figs \@ref(fig:biomass), \@ref(fig:taxa), and \@ref(fig:traits)... -->
It was initially hypothesized that community composition would vary with distance to edge, favoring more shade-tolerant taxa, due to light availability and dispersal potential, 
and the evidence presented here supported edge effects on community composition and biomass of taxa associated with both early- and late-successional stages, but without detectable roles for light availability or general dispersal mode.
Accordingly, this suggests that edge effects do significantly change humid tropical forest community taxonomic and functional composition, but primarily due to factors other than light competition or dispersal limitation *[@krishnadas20]*, and also suggests that aboveground biomass and even low taxonomic richness are resilient after a decade of regeneration.

Aboveground biomass storage is a key global ecosystem function and service, but edge effects tend to release about one-third of carbon lost from tropical deforestation *[@brinck17]*, with more expected in the future *[@mitchard18]*.
<!-- responds variably to secondary succession and edge effects ...   -->
Management has been found to explain most of biomass variation among tropical forest fragments, and wood density ~10% biomass variation *[@pyles22]*.
This study adds that management via edge effects can also mediate ~10% of wood density variation without affecting local biomass, although median stem wood density can describe regional spatial biomass patterns *[@baker04]*.
In this study, wood density and richness values may have been pulled down by early-successional clonal trees closest to the primary forest, which tend to have lower wood densities, lowering carbon *[@gonzalezj98; @pyles22]*, and increasing nearby stem density.
Lower wood density can also reflect more dynamic forest stands *[@malhi06]* with higher tree turnover, a consequence of forest fragmentation *[@nascimento04a]* together with other global change drivers *[@laurance14]*, as well as may reflect possible differences in soil fertility *[@malhi04]*, all of which represent possible future research directions, beyond strictly neutral dynamics *[@terborgh95]*.
Furthermore, light availability at the plot level did not affect biomass storage, but instead, individual light gaps may be more important for biomass dynamics *[@chazdon84]*.
    <!-- - although canopy light may still affect biomass generally in this case, without edge effects (check direct light-biomass correlation graph? -- no sig correlation) -->

Edge effects have been shown in other studies to lower both forest biomass and change community composition *[@anderson22]*, and this study adds that the decline in diversity can be rapid and non-linear across edge gradients, along with supporting the likelihood of lasting changes to community composition.
This rapid loss of diversity across the edge gradient may be in part due to existing higher yet constant levels of shade limiting shade-intolerant seedling recruitment and/or performance, together with insufficient time for slower-growing shade-tolerant taxa to accumulate significant amounts of biomass.
Another possible factor may be overall short dispersal kernels, in part due to higher fungal pathogen pressure under higher constant shade, also increasing tree mortality.
Decomposing compositional changes into key specific taxa, while the biomass of the most abundant taxon *Vochysia* was resilient to edge effects in this study, that of the key taxon for frugivores like birds and bats *Ficus* was significantly lowered across the secondary forest edge, pointing to potentially different responses of biomass and food web associated ecosystem services.
Community composition also tended to show lower variance across the edge gradient, although this effect cannot be confirmed here given the uneven sampling that accounted for variable forest area at different distances from edge.

Characterizing community composition by functional traits is also increasingly studied *[@kearney21]*, but less so regarding edge effects.
A study in Madagascar found lower phylogenetic diversity closer to degraded forest edge, but no change in aboveground biomass *[@razafindratsima18]*.
These results parallel in showing lower taxonomic diversity, although biomass specifically of taxa assoociated with both early- and late-successional stages also decreased with away from primary forest edge.
This result by successional stage can be explained by slightly lower habitat quality inferred near exposed habitat edges near roads, and accordingly relatively higher habitat quality near closed primary forest edge.
While lower biomass of succession-agnostic taxa further from primary forest edge appears consistent with dispersal kernels, the lack of pattern by broad dispersal mode limits this explanation here, yet could be further tested by measuring taxon-specific dispersal kernels.

Overall, this study highlights how abandoned wet tropical timber plantation can regenerate in alongside fragmented forests.
Results support related syntheses that biomass and taxonomic richness often regrows relatively quickly, 
<!-- **along with traits?,** -->
while taxonomic composition recovers much more slowly, if not diverging altogether *[@norden15]*.
These findings can suggest that more efficient reforestation could more actively target biodiversity conservation over other resilient functions like carbon storage, and that restoration efforts could include focusing activities like native species planting efforts on degraded habitat edges away from primary forest edges, and investing less in areas closer to existing forest edges, where recruitment may already maintain diversity levels.
As forests continue to become fragmented *[@haddad15]*, understanding how to work with natural regeneration patterns around edge effects will likely become increasingly important.

### 2022-07-20 article/intro.Rmd
start draft disc
first poster draft
ideas ATBC
draft disc

Forest and landscape restoration is a key international conservation and climate change adaptation strategy *[@depinto20]*, and while tropical forests specifically store most land biomass *[@pan13]*, most are now secondary *[@fao20]* and regenerating degraded functions *[@hubau20]* amid increasing climate stressors *[@anderegg22]* yet insufficient policy support *[@chazdon18]* via skewed priorities *[@pyron22]*.
Secondary forests can regenerate relatively quickly compared to old-growth forests in some ways, but in other ways remain distinct.
Neotropical secondary forests can grow quickly enough to accumulate biomass >10x faster than old-growth forests *[@finegan96; @poorter16]*, and substantially offset carbon emissions, with estimates ranging from ~10% from Amazonian deforestation *[@smith20]* to a decade of fossil fuel emissions from all of Latin American and the Caribbean *[@chazdon16]*.
They also recover taxonomic richness and biomass quite quickly, as well as tree height, averaging 80% of old-growth forest levels in just 20 years, especially in wetter regions *[@rozendaal19]*.
The recovery of species richness alone is beneficial in that it also tends to correlate with the recovery of some ecosystem services *[@guariguata01; @stanturf14]* like biomass and carbon storage *[@liu18]*.
However, secondary neotropical forests diverge widely in taxonomic composition *[@norden15]*, potentially taking over a century to recover *[@poorter21]* with added variability *[@atkinson22]*.
Despite being one of the slowest ecosystem properties to recover, restoring recovering community composition is often important for preserving rare taxa *[@carlo16]*, which can be keystone to locally-adapted food webs, including birds *[@maas16]*.
As a result, understanding secondary forest regeneration remains important...

Secondary forest regeneration is often affected by edge effects, or changes near an ecosystem boundary, as well as land-use history, especially in heterogeneously-managed landscapes *[@perfecto09; @melo13]*.
About 70% of all forests have been estimated to be within a km of their edge *[@haddad15]*, resulting in tropical forests having been estimated to lose ~22% aboveground carbon along heavily-managed edges >100m, in part due to shortening trees and lighter leaf mass *[@ordway20]*.
Edge effects also have the potential to shape community composition, based on the local dispersal kernel patterns of existing taxa *[@wandrag17]*, which while complicated *[@muller-landau09]*, can also correlate with simpler plant traits like height and seed mass *[@thomson11; @tamme14]*.
The traits of existing taxa also depend on land-use history and previous management *[@omeja12]*, which historically has favored tree taxa that tend to pioneering, or be faster-growing, have lower wood densities, and be less shade-tolerant *[@tabarelli08]*.
Legacy effects from management like intensive timber planting could in theory delay or result in non-linear changes between successional stages *[@albrich20; @gough21]*, which would make future management more uncertain, unlike more sustainable thinning practices and agroforestry *[@lefland18]*.
Overall, while edge effects near degraded habitat edges are often negative, meta-population *[@levins79]* and -community *[@leibold04; @warren15]* theories highlight dispersal as a key process for offsetting extinction debt and thus maintaining biodiversity.
Near primary forest, dispersal may facilitate secondary forest regeneration, and help explain relatively high biodiversity compared to adjacent stands that can be found in agroforests *[@oliveira-neto17]*, forests affected by logging *[@clark12; @edwards14]*, and timber plantations *[@pryde15]*.
<!-- role of N-fixing trees... *[@lai18]* -->

@@

>>>>

Forest and landscape restoration is a key international conservation and climate change adaptation strategy *[@depinto20]*, and while tropical forests specifically store most land biomass *[@pan13]*, most are now secondary *[@fao20]* and regenerating degraded functions *[@hubau20]* amid increasing climate stressors *[@anderegg22]* yet insufficient policy support *[@chazdon18]* via skewed priorities *[@pyron22]*.
Secondary forests can regenerate relatively quickly compared to old-growth forests in some ways, but in other ways remain distinct.
Neotropical secondary forests can grow quickly enough to accumulate biomass >10x faster than old-growth forests *[@finegan96; @poorter16]*, and substantially offset carbon emissions, with estimates ranging from ~10% from Amazonian deforestation *[@smith20]* to a decade of fossil fuel emissions from all of Latin American and the Caribbean *[@chazdon16]*.
They also recover taxonomic richness and biomass quite quickly, as well as tree height, averaging 80% of old-growth forest levels in just 20 years, especially in wetter regions *[@rozendaal19]*.
The recovery of species richness alone is beneficial in that it also tends to correlate with the recovery of some ecosystem services *[@guariguata01; @stanturf14]* like biomass and carbon storage *[@liu18]*.
However, secondary neotropical forests diverge widely in taxonomic composition *[@norden15]*, potentially taking over a century to recover *[@poorter21]* with added variability *[@atkinson22]*.
Despite being one of the slowest ecosystem properties to recover, restoring recovering community composition is often important for preserving rare taxa *[@carlo16]*, which can be keystone to locally-adapted food webs, including birds *[@maas16]*.
Understanding secondary forest regeneration is key for global biodiversity conservation at the global scale, yet local management remains the strongest driver of tropical forest carbon stores *[@pyles22]*.

Secondary forest regeneration is often affected by edge effects, or changes near an ecosystem boundary, as well as land-use history, especially in heterogeneously-managed landscapes *[@perfecto09; @melo13]*.
About 70% of all forests have been estimated to be within a km of their edge *[@haddad15]*, resulting in tropical forests having been estimated to lose ~22% aboveground carbon along heavily-managed edges >100m, in part due to shortening trees and lighter leaf mass *[@ordway20]*.
Edge effects also have the potential to shape community composition, based on the local dispersal kernel patterns of existing taxa *[@wandrag17]*, which while complicated *[@muller-landau09]*, can also correlate with simpler plant traits like height and seed mass *[@thomson11; @tamme14]*.

Properties of the existing tree community also depend on land-use history and previous management *[@omeja12]* including plantation use for timber production, which historically has favored tree taxa that tend to pioneering, or be faster-growing, have lower wood densities, and be less shade-tolerant *[@tabarelli08]*.
Legacy effects from management like intensive timber planting could in theory delay or result in non-linear changes between successional stages *[@albrich20; @gough21]*, which would make future management more uncertain, unlike more sustainable thinning practices and agroforestry *[@lefland18]*.
Overall, while edge effects near degraded habitat edges are often negative, meta-population *[@levins79]* and -community *[@leibold04; @warren15]* theories highlight dispersal as a key process for offsetting extinction debt and thus maintaining biodiversity.
Near primary forest, dispersal may facilitate secondary forest regeneration, and help explain relatively high biodiversity compared to adjacent stands that can be found in agroforests *[@oliveira-neto17]*, forests affected by logging *[@clark12; @edwards14]*, and timber plantations *[@pryde15]*.

<!-- role of N-fixing trees... *[@lai18]* -->
<!-- A subset of secondary forests include naturally regenerating timber plantations, which may show unique recovery patterns. -->
<!-- .. -->
<!-- More specifically, regenerating timer plantations may also experience edge effects, but they tend to be ...  -->
<!-- - from *[@anderson22; others]* -->
<!-- - examples of *B. quinata* plantation papers specifically -->

@@

<!-- - conceptual hypothesis diagram? -->

### 2022-07-20 article/merge.Rmd
first poster draft
ideas ATBC
draft disc

---
title: ""
author: 
  name: Nicholas Medina
  affilitation: Brandeis University
address:
  code: Brandeis University
  address: 415 South St, Waltham, MA 02453
output: 
  # html_notebook
  bookdown::html_document2:
    number_sections: FALSE
bibliography: ../../library.bib
link-citations: TRUE
---

# Abstract

@@

@@

<!-- # ```{r child="results.Rmd"} -->
<!-- # ``` -->

@@

>>>>

---
title: "Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest"
author: 
  - Nicholas Medina ^a^*
  - Max Villalobos ^b^
  - Andrea Johnson ^b^
date: 
  scriptsize\ ^a^ Biology and Environmental Studies, Brandeis University, 415 South St, Waltham, MA, USA 02453;
  ^b^ Land Stewardship, Osa Conservation, Puerto Jiménez, Golfito, Costa Rica
  # Ecology and Evolutionary Biology, University of Michigan, 1105 N University Ave, Ann Arbor, MI, USA 48109
# author: 
#   - name: "Nicholas Medina"
#     email: nmedina@umich.edu
#     affiliation: Brandeis University
#     orcid_id: 0000-0001-5465-3988
# address:
#   - code: Brandeis University
#     address: Biology and Environmental Studies, 415 South St, Waltham, MA 02453
# keywords: [1, 2, 3]

output: 
  # html_notebook
  # bookdown::html_document2:
  #   # df_print: paged
  #   global_numbering: TRUE
  #   toc: FALSE
  bookdown::pdf_document2:
    # number_sections: FALSE
    global_numbering: TRUE
    keep_tex: TRUE
    toc: FALSE
bibliography: 
  - ../../library.bib
  - ../../grateful-refs.bib
  - ../../software.bib
link-citations: TRUE
csl: ../../styles-master/elsevier-harvard.csl

fontsize: 12pt
geometry: margin=1in
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{setspace}\doublespace
# abstract
---

<!-- journals: J [Appl] Ecol, J [Appl] Veg Sci, Basic Appl Ecol (3.5), Biotropica (2.5), Eco Evo -->

\
\

**Keywords**: secondary forest; regeneration; timber plantation associated biodiversity; edge effect; aboveground biomass; shade

\
\

***Corresponding author**: Nicholas Medina, nmedina@umich.edu

\newpage

# Abstract{-}

@@

\newpage

@@

```{r child="results.Rmd"}
```

@@

\newpage

# Funding{-}

This study was funded by Brandeis University Hiatt Career Center World of Work Fellowship.

# Declaration of interests{-}

Authors declare no conflicts of interest.

# Acknowledgements{-}

Thanks to Professor Dan Perlman for initial networking; 
<!-- concurrent Osa Conservation managers Max Villalobos and Andrea Johnson for project guidance and supervision;  -->
Lehigh University Professor Don Morris and interns Emma Capetz and Erin Lau along with field station staff Agustín Mendoza for direct field assistance; 
concurrent Osa Conservation general manager Manuel Ramirez and Piro Biological Station staff Juan Carlos Cruz Díaz, Annia Barrantes, Larry Villalobos, Hansel Vargas, and visitors for field support; 
peer James Fifer for discussion of early draft; 
and Luisa Valdez for translation assistance.

# Author contributions{-}

MV and AJ contributed to study conceptualization, administration, and supervision; NM and AJ contributed to funding acquisition; NM contributed to data collection; NM and MV contributed to data analysis;
NM wrote initial draft and later revisions.

# Data statement{-}

Code stored at [github.com/nmedina17/osa](github.com/nmedin17/osa) *[@softwareMedina22osa]*.

\newpage
